---
title: "Use case 3: CytoNorm_channels_without_clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Use case 3: CytoNorm_channels_without_clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- github markdown built using
rmarkdown::render("vignettes/CytoNorm_channels_without_clustering.Rmd", output_format = "github_document")
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#### In this vignette, you learn how CytoNorm can be used to normalize markers that are not used during the clustering step.

One of the strengths of CytoNorm and some other normalization algorithms is the clustering step. By clustering first, cell subsets are separated, which makes itis possible to apply different normalizations on the different subsets. However,in cytometry, we often make the distinction between phenotypic or cell type markers and functional or cell state markers and clustering is typically done only with the phenotypic markers, as functional markers typically do not contribute to the clustering, or even reduce the clustering quality. Therefore, it is important that the normalization algorithm allows making a distinctionbetween the channels that need to be normalized, which might include functional markers, and the channels that will be used for the clustering. 

By default, CytoNorm uses all the markers specified for normalization also for clustering. It is however straightforward to specify different sets of markers for the clustering step (in the FlowSOM parameters) and for the actualnormalization procedure.

## Prepare the CytoNorm normalization step
The phenotypic markers of our mesothelioma dataset were already normalized with the two models in the previous use cases ([Use case 1: Run CytoNorm on a dataset without dedicated controls vignette](CytoNorm_without_controls.md) and [Use case 2: Run CytoNorm to normalize towards a distribution vignette](CytoNorm_towards_distribution.md). Since we want to further characterize the immune landscape, the next step was is analyze the functional markers. A manual gating strategy of these markers was designed and optimized by an expert for the first cohort of data (P1_C1 and P2_C1), which should be translated to the second cohort (P1_C2 and P2_C2). 

To do this, we train a CytoNorm model per panel where we set the goal to the first cohort (P1_C1 or P2_C1), use the cell type markers for the FlowSOMclustering and specify which (panel-specific) cell state markers should be normalized.

```{r setup}
library(CytoNorm)
library(flowCore)
library(FlowSOM)
library(ggpubr)
```

## Train the CytoNorm model
### List the files per batch that will be aggregated
```{r}
dir_prepr <- "Data/Preprocessed"

# Organize preprocessed data in batches
files_P1_C2 <- list.files(path = "Data/Preprocessed/",
                          pattern = "ID[5-8]_Panel1_TP[1-3].fcs",
                          full.names = TRUE)

# List files (already normalized for cell type markers in use cases 1 and 2)
files_P1_C1_norm <- list.files(path = "Data/Normalized_cellType", 
                               pattern = "ID[1-4]_Panel1_TP..fcs", full.names = TRUE)
files_P1_C2_norm <- list.files(path = "Data/Normalized_cellType", 
                               pattern = "ID[5-8]_Panel1_TP..fcs", full.names = TRUE)

# Read in 1 file per panel to get channel information
ff_P1 <- read.FCS(files_P1_C1_norm[1])

# Retrieve channel information
cellTypeMarkers <- c("CD27", "CD38", "IgD", "IgM") 
cellTypeChannels <- GetChannels(object = ff_P1, markers = cellTypeMarkers)
P1_cellStateMarkers <- c("CD5", "CD268", "CD24", "PDL1", "PD1", "CD86", "KI67")
P1_cellStateChannels <- GetChannels(object = ff_P1, markers = P1_cellStateMarkers)

# Read in manualLabels object that was generated in the CytoNorm_without_controls vignette
manualLabels <- readRDS("Data/Preprocessed/attachments/manualLabels.rds")
```

### Train the CytoNorm model
```{r Train CytoNorm}
model3_cellStateMarkersP1 <- CytoNorm.train(files = c(files_P1_C1_norm, files_P1_C2_norm),
                                            labels = c(rep(x = "C1",
                                                           times = length(files_P1_C1_norm)),
                                                       rep(x = "C2",
                                                           times = length(files_P1_C2_norm))),
                                            channels = P1_cellStateChannels,
                                            transformList = NULL,
                                            seed = 1,
                                            plot = TRUE,
                                            verbose = TRUE,
                                            normParams = list("goal" = "C1"),
                                            FlowSOM.params = list(nCells = 1e+06, 
                                                                  xdim = 7, ydim = 7, 
                                                                  nClus = 3, 
                                                                  scale = FALSE,
                                                                  colsToUse = cellTypeChannels))
saveRDS(object = model3_cellStateMarkersP1, file = "RDS/model3_cellStateMarkersP1.rds")
```

By default, the FlowSOM clustering will be done with the `channels` listed in the `CytoNorm.train()` call. However, it is possible to provide other channels for the FlowSOM clustering by listing them at the `colsToUse` argument within the `FlowSOM.params`. Note that in the workflow of this paper, CytoNorm will again reuse the previous FlowSOM model, as discussed before.

## Apply the CytoNorm model
```{r Apply CytoNorm}
CytoNorm.normalize(model = model3_cellStateMarkersP1,
                   files = c(files_P1_C1_norm, files_P1_C2_norm),
                   labels = c(rep(x = "C1",times = length(files_P1_C1_norm)),
                              rep(x = "C2",times = length(files_P1_C2_norm))),
                   transformList = NULL,
                   verbose = TRUE,
                   prefix = "",
                   transformList.reverse = NULL, 
                   outputDir = "Data/Normalized_cellState")

```

## Evaluate the quality
### Density plots
```{r Evaluate densities, fig.height=28, fig.width=6}
# List files
files_P1_C1_norm_norm <- list.files(path = "Data/Normalized_cellState", 
                                    pattern = "ID[1-4]_Panel1_TP..fcs", full.names = TRUE)
files_P1_C2_norm_norm <- list.files(path = "Data/Normalized_cellState", 
                                    pattern = "ID[5-8]_Panel1_TP..fcs", full.names = TRUE)

# Make plots
p <- plotDensities(input = list("P1_C1" = files_P1_C1_norm,
                                "P1_C2" = files_P1_C2_norm,
                                "P1_C1_norm" = files_P1_C1_norm_norm,
                                "P1_C2_norm" = files_P1_C2_norm_norm),
                   channels = P1_cellStateChannels,
                   colors = c("P1_C1" = "#900002","P1_C2" = "#FF3437"), 
                   model = model3_cellStateMarkersP1)

# Arrange figure
p_ <- ggarrange(ggarrange(plotlist = p[1:length(p)-1], 
                          ncol = (length(p)-1)/(2*length(P1_cellStateChannels)), 
                          nrow = 2*length(P1_cellStateChannels)),
                widths = c(1,3),
                p$legend, nrow = 2, heights = c(10,1))
print(p_)
```

### Spline plots
```{r Evaluate splines, fig.height=9, fig.width=12}
plotSplines(model = model3_cellStateMarkersP1, channels = P1_cellStateChannels, groupClusters = TRUE)
```

### EMD and MAD
#### Make aggregates and get aggregate manual labels
```{r Get manual labels}
aggregates <- list("agg_P1_C2" = files_P1_C2)

for (agg in names(aggregates)){
  writeLines(agg)
  files <- aggregates[[agg]]
  set.seed(2023)
  before <- AggregateFlowFrames(fileNames = files, silent = TRUE,
                                cTotal = length(files) * 10000)

  write.FCS(before, paste0("tmp/before/", agg, ".fcs"))  
  after <- before
  
  manual <- c()
  for (i_file in unique(before@exprs[,"File"])){
    ff <- read.FCS(paste0("Data/Normalized_cellType/", basename(aggregates[[agg]][i_file])))
    i_IDs <- before@exprs[before@exprs[,"File"] == i_file,"Original_ID2"]
    after@exprs[before@exprs[,"File"] == i_file,1:(ncol(after@exprs)-3)] <- ff@exprs[i_IDs, ]
    labels <- as.character(manualLabels[[basename(aggregates[[agg]][i_file])]])
    manual <- c(manual, labels[i_IDs])
  }    
  
  manualLabels[[paste0(agg, ".fcs")]] <- factor(manual, levels = levels(manualLabels[[1]]))
  write.FCS(after, paste0("tmp/after/", agg, ".fcs"))
}
```

#### Define cell types and files for model evaluation
```{r Specify cell types of interest}
cellTypes <- c("unlabeled",
               "All B cells",
               "Transitional B cells",
               "Naive B cells",
               "IgD+ memory B cells",
               "IgM+ memory B cells", 
               "Class switched memory B cells", 
               "Double negative B cells")

models <- list("Model3a_before" = list("files" = c(files_P1_C1_norm, "tmp/before/agg_P1_C1.fcs", 
                                                    "tmp/before/agg_P1_C2.fcs", files_P1_C2_norm),
                                       "channels" = P1_cellStateChannels),
               "Model3a_after" = list("files" = c(files_P1_C1_norm_norm, "tmp/after/agg_P1_C1.fcs", 
                                                   "tmp/after/agg_P1_C2.fcs", files_P1_C2_norm_norm),
                                      "channels" = P1_cellStateChannels))
```

#### Calculate EMD and MAD
```{r Calculate EMD and MAD}
EMDs <- list()
MADs <- list()

for (model in names(models)){
  print(model)
  files <- models[[model]][["files"]]
  channels <- models[[model]][["channels"]]
  
  EMDs[[model]] <- emdEvaluation(files = files,
                                 channels = channels,
                                 manual = manualLabels, return_all = TRUE, 
                                 manualThreshold = 20)
  MADs[[model]] <- CytoNorm:::madEvaluation(files = files[!startsWith(files, "tmp")], 
                                 channels = channels, return_all = TRUE,
                                 manual = manualLabels, transformList = NULL)
}

EMD_scores <- list()
for (subset in names(models)){
  dist <- EMDs[[subset]]$distances
  EMD_scores[[subset]] <- matrix(NA,
                                 nrow = length(dist), ncol = length(dist[[1]]),
                                 dimnames = list(names(dist),
                                                 names(dist[[1]])))
  for (cellType in names(dist)){
      for (marker in names(dist[[cellType]])){
        EMD_scores[[subset]][cellType, marker] <- dist[[cellType]][[marker]][13, 14]
      }
  }
}
```

#### Make figures
```{r EMD and MAD figures, fig.height=5, fig.width=12}
markerlevels <- NULL
plotlist <- list()

# EMD plot
EMD_before <- EMD_scores[["Model3a_before"]][-2,]
EMD_after <- EMD_scores[["Model3a_after"]][-2,]

EMD_df <- data.frame("before" = c(EMD_before),
                     "after" = c(EMD_after),
                     "feature" = paste(rownames(EMD_before), 
                                       rep(colnames(EMD_before), 
                                           each = nrow(EMD_before)), sep = "_"))
if (is.null(markerlevels)){
  markerlevels <- unique(unique(sub(".*_", "", EMD_df$feature)))
}
EMD_df$marker <- factor(sub(".*_", "", EMD_df$feature), levels = markerlevels)
EMD_df$celltype <- factor(sub("_.*", "", EMD_df$feature), levels = c("AllCells", cellTypes))
max <- max(EMD_df[,1:2], na.rm = T)
p_emd <- ggplot(EMD_df, aes(x = after, y = before)) +
  xlim(0,max) + ylim(0,max) +
  geom_point(aes(color = marker, shape = celltype)) +
  geom_abline() +
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 6)) +
  ggtitle("EMD") +
  theme_minimal()

leg_emd <- get_legend(p_emd)

p_emd <- p_emd  +
    theme(legend.position = "none")
plotlist[[length(plotlist)+1]] <- p_emd    

names(markerlevels) <- sub(".*<(.*)>.*", "\\1", markerlevels)

# MAD plot
MAD_before <- MADs[["Model3a_before"]][["comparison"]][-2,]
MAD_after <- MADs[["Model3a_after"]][["comparison"]][-2,]
MAD_df <- data.frame("before" = c(MAD_before),
                     "after" = c(MAD_after),
                     "feature" = paste(rownames(MAD_before), 
                                       rep(colnames(MAD_before), 
                                           each = nrow(MAD_before)), sep = "_"))
MAD_df$marker <- factor(markerlevels[sub(".*_", "", MAD_df$feature)], levels = markerlevels)
MAD_df$celltype <- factor(sub("_.*", "", MAD_df$feature), c("AllCells", cellTypes))
max <- max(MAD_df[,1:2], na.rm = T)
p_mad <- ggplot(MAD_df, aes(x = after, y = before)) +
  xlim(0,max) + ylim(0,max) +
  geom_point(aes(color = marker, shape = celltype)) +
  geom_abline() +
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8, 6)) +
  ggtitle("MAD") +
  theme_minimal()

leg_mad <- get_legend(p_mad)

p_mad <- p_mad  +
    theme(legend.position = "none")

plotlist[[length(plotlist)+1]] <- p_mad

ggarrange(ggarrange(plotlist = plotlist),
          as_ggplot(leg_emd), ncol=2, widths = c(4,1.5))
```

